Assessing the spatial patterning of Middle Paleolithic human settlement in Western Iberia

### 2020-10-21 analysis/paper/paper.Rmd
most statistics into RMarkdown

---
title: "Assessing spatial patterning of Middle Paleolithic human settlement in Western Iberia"
author:
  - João Cascalheira
  - Célia Gonçalves
  - Daniela Maio
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  Middle Paleolithic; Settlement patterns; GIS; Western Iberia
---

@@

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

@@

library(rgdal)
library(raster)
library(rasterVis)
library(lattice)
library(foreach)
library(rnaturalearth)
library(tidyverse)
library(ggplot2)
library(broom)
library(foreach)
library(sf)

@@

#Download geo tif from OSF repositorium

#osf_retrieve_file("nv6r8") %>%
#  osf_download(path = "./analysis/data/raw_data")
#osf_retrieve_file("jb6v8") %>%
#  osf_download(path = "./analysis/data/raw_data")
#osf_retrieve_file("yxk23") %>%
#  osf_download(path = "./analysis/data/raw_data")
#osf_retrieve_file("57qfn") %>%
#  osf_download(path = "./analysis/data/raw_data")
#osf_retrieve_file("mx2bt") %>%
#  osf_download(path = "./analysis/data/raw_data")

@@

#Import sites
sites <- read.csv("../data/raw_data/sites_short_list.csv", header=TRUE)
caves <- read.csv("../data/raw_data/all_caves.csv", header=TRUE)

#Convert to shapefile
coordinates(sites)<-~Longitude+Latitude
proj4string(sites) = CRS("+init=epsg:4326")
writeOGR(sites, "../data/raw_data/spatial_data/shapefiles","sites_short", "ESRI Shapefile")

coordinates(caves)<-~Longitude+Latitude
proj4string(caves) = CRS("+init=epsg:4326")
writeOGR(caves, "../data/raw_data/spatial_data/shapefiles","caves", "ESRI Shapefile")

@@

# Subdivide sites for further analysis
open_air <- dplyr::filter(as.data.frame(sites_vals), Type == "Open air")
cave <- dplyr::filter(as.data.frame(sites_vals), Type == "Cave")

@@

@@

# Plot distributions
ggplot() +
  geom_line(data = background_altitude_densities,
            mapping = aes(x = Elevation,
                          y = Frequency)) +
  geom_ribbon(data = background_altitude_densities,
              mapping = aes(x = Elevation,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_altitude_densities,
            mapping = aes(x = Elevation,
                          y = Frequency),
            color = "red")

@@

# Plot distributions
ggplot() +
  geom_line(data = background_slope_densities,
            mapping = aes(x = Slope,
                          y = Frequency)) +
  geom_ribbon(data = background_slope_densities,
              mapping = aes(x = Slope,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_slope_densities,
            mapping = aes(x = Slope,
                          y = Frequency),
            color = "red")

@@

# Plot distributions
ggplot() +
  geom_line(data = background_aspect_densities,
            mapping = aes(x = Aspect,
                          y = Frequency)) +
  geom_ribbon(data = background_aspect_densities,
              mapping = aes(x = Aspect,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_aspect_densities,
            mapping = aes(x = Aspect,
                          y = Frequency),
            color = "red")

@@

# Plot distributions
ggplot() +
  geom_line(data = background_distance_densities,
            mapping = aes(x = Distance,
                          y = Frequency)) +
  geom_ribbon(data = background_distance_densities,
              mapping = aes(x = Distance,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_distance_densities,
            mapping = aes(x = Distance,
                          y = Frequency),
            color = "red")

@@

>>>>

---
title: "Assessing the spatial patterning of Middle Paleolithic human settlement in Western Iberia"
author:
  - João Cascalheira
  - Célia Gonçalves
  - Daniela Maio
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  Middle Paleolithic; Settlement patterns; GIS; Western Iberia
---

@@

```{r, setup, eval = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

@@

library(rgdal)
library(raster)
library(rasterVis)
library(lattice)
library(foreach)
library(rnaturalearth)
library(tidyverse)
library(ggplot2)
library(broom)
library(foreach)
library(sf)
library(osfr)
library(cowplot)
library(gridExtra)
library(rela)
library(psych)
library(factoextra)

@@

#Download geo tiffs from OSF repositorium when running this document for the first time

#osf_retrieve_file("nv6r8") %>%
#  osf_download(path = "../data/raw_data")
#osf_retrieve_file("jb6v8") %>%
#  osf_download(path = "../data/raw_data")
#osf_retrieve_file("yxk23") %>%
#  osf_download(path = "../data/raw_data")
#osf_retrieve_file("57qfn") %>%
#  osf_download(path = "../data/raw_data")
#osf_retrieve_file("mx2bt") %>%
#  osf_download(path = "../data/raw_data")

@@

#Import sites and convert to shapefile
#sites <- read.csv("../data/raw_data/sites_short_list.csv", header=TRUE)
#caves <- read.csv("../data/raw_data/all_caves.csv", header=TRUE)

#Convert to shapefile
#coordinates(sites)<-~Longitude+Latitude
#proj4string(sites) = CRS("+init=epsg:4326")
#writeOGR(sites, "../data/raw_data/spatial_data/shapefiles","sites_short", "ESRI Shapefile")

#coordinates(caves)<-~Longitude+Latitude
#proj4string(caves) = CRS("+init=epsg:4326")
#writeOGR(caves, "../data/raw_data/spatial_data/shapefiles","caves", "ESRI Shapefile")

@@

# Subdivide sites for further analysis
open_air <- dplyr::filter(as.data.frame(sites_vals), Type == "Open air")
caves <- dplyr::filter(as.data.frame(sites_vals), Type == "Cave")
all_caves <- as.data.frame(caves_vals)

@@

## Open air

@@

# Plot distributions
open_air_elevation <- ggplot() +
  geom_line(data = background_altitude_densities,
            mapping = aes(x = Elevation,
                          y = Frequency)) +
  geom_ribbon(data = background_altitude_densities,
              mapping = aes(x = Elevation,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_altitude_densities,
            mapping = aes(x = Elevation,
                          y = Frequency),
            color = "red") +
  theme_cowplot() +
  ylab("") + xlab("meters")

@@

# Plot distributions
open_air_slope <- ggplot() +
  geom_line(data = background_slope_densities,
            mapping = aes(x = Slope,
                          y = Frequency)) +
  geom_ribbon(data = background_slope_densities,
              mapping = aes(x = Slope,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_slope_densities,
            mapping = aes(x = Slope,
                          y = Frequency),
            color = "red") +
  theme_cowplot() +
  ylab("") + xlab("degrees")

@@

# Plot distributions
open_air_aspect <- ggplot() +
  geom_line(data = background_aspect_densities,
            mapping = aes(x = Aspect,
                          y = Frequency)) +
  geom_ribbon(data = background_aspect_densities,
              mapping = aes(x = Aspect,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_aspect_densities,
            mapping = aes(x = Aspect,
                          y = Frequency),
            color = "red") +
  theme_cowplot() +
  ylab("") + xlab("degrees")

@@

# Plot distributions
open_air_distance <- ggplot() +
  geom_line(data = background_distance_densities,
            mapping = aes(x = Distance,
                          y = Frequency)) +
  geom_ribbon(data = background_distance_densities,
              mapping = aes(x = Distance,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_distance_densities,
            mapping = aes(x = Distance,
                          y = Frequency),
            color = "red") +
  theme_cowplot() +
  ylab("") + xlab("meters")

@@

```{r open-air-plots, eval = FALSE, echo = FALSE}

plot_grid(open_air_elevation, open_air_slope, open_air_aspect, open_air_distance, labels = c("A", "B", "C", "D"), nrow = 2, ncol = 2)

## Caves

```{r caves-boxplots, eval = FALSE, echo = FALSE}

cave <- caves %>% 
  dplyr::select(SITE, areaDEM_cut, areaSLOPE_cut, areaASPECT_cut, areaDIST_cut2) %>% 
  mutate(TYPE = "MP Caves")

all_caves <- all_caves %>% 
  rename(SITE = Ã.__SITE) %>% 
  dplyr::select(SITE, areaDEM_cut, areaSLOPE_cut, areaASPECT_cut, areaDIST_cut2) %>% 
  mutate(TYPE = "All Caves")

cave <- rbind(cave, all_caves)

caves_elevation <- cave %>%
  mutate(TYPE = fct_relevel(TYPE, 
            "MP Caves", "All Caves")) %>% 
  ggplot(aes(x = TYPE, y = areaDEM_cut, fill = TYPE)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  ylab("meters") + xlab("") +
  theme_cowplot() +
  theme(legend.position="none") +
  scale_fill_brewer(palette = "Dark2")

caves_slope <- cave %>%
  mutate(TYPE = fct_relevel(TYPE, 
            "MP Caves", "All Caves")) %>%
  ggplot(aes(x = TYPE, y = areaSLOPE_cut, fill = TYPE)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  ylab("degrees") + xlab("") +
  theme_cowplot() +
  theme(legend.position="none") +
  scale_fill_brewer(palette = "Dark2")

caves_aspect <- cave %>% 
  mutate(TYPE = fct_relevel(TYPE, 
            "MP Caves", "All Caves")) %>%
  ggplot(aes(x = TYPE, y = areaASPECT_cut, fill = TYPE)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  ylab("degrees") + xlab("") +
  theme_cowplot() +
  theme(legend.position="none") +
  scale_fill_brewer(palette = "Dark2")

caves_distance <- cave %>% 
  mutate(TYPE = fct_relevel(TYPE, 
            "MP Caves", "All Caves")) %>%
  ggplot(aes(x = TYPE, y = areaDIST_cut2, fill = TYPE)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  ylab("meters") + xlab("") +
  theme_cowplot() +
  theme(legend.position="none") +
  scale_fill_brewer(palette = "Dark2")

plot_grid(caves_elevation, caves_slope, caves_aspect, caves_distance, labels=c("A", "B", "C", "D"), ncol = 2, nrow = 2)

```{r caves-statistics, eval = FALSE, echo = FALSE}

caves_elevation_MWU <- wilcox.test(areaDEM_cut ~ TYPE, data = cave,
                   exact = FALSE)

caves_slope_MWU <- wilcox.test(areaSLOPE_cut ~ TYPE, data = cave,
                   exact = FALSE)

caves_aspect_MWU <- wilcox.test(areaASPECT_cut ~ TYPE, data = cave,
                   exact = FALSE)

caves_distance_MWU <- wilcox.test(areaDIST_cut2 ~ TYPE, data = cave,
                   exact = FALSE)

variables <- c("Elevation", "Slope", "Altitude", "Distance")

u_statistics <- c(caves_elevation_MWU$statistic, caves_slope_MWU$statistic,
                   caves_aspect_MWU$statistic, caves_distance_MWU$statistic)

p_values <- c(caves_elevation_MWU$p.value, caves_slope_MWU$p.value,
                   caves_aspect_MWU$p.value, caves_distance_MWU$p.value)

caves_stats <- data.frame(variables, u_statistics, p_values)

caves_stats <- caves_stats %>% 
  rename(Variables = variables, `U statistic` = u_statistics, "p-value" = p_values) %>% 
  mutate(across(where(is.numeric), round, 2))

knitr::kable(caves_stats, caption = "Results of the Mann-Whitney U test, for each spatial variable, between caves with Middle Paleolithic occupations and all other archaeological caves identified in westernmost Iberia.")

## Principal Component Analysis

```{r PCA, eval = FALSE, echo = FALSE}

pca_data <- as.data.frame(sites_vals)

pca_data <- pca_data %>% 
  dplyr::select(SITE, Type, Dmn_R_M, Elevation = areaDEM_cut, Slope = areaSLOPE_cut, Aspect = areaASPECT_cut,
                Distance = areaDIST_cut2)

pca_data_active <- dplyr::select(pca_data, Elevation,
                                 Aspect, Slope, Distance)

# Test assumptions
pca_Cor <- cor(pca_data_active)

# Sampling adequacy or an appropriate number of observations relative to the number of variables being examined
assumptions <- paf(as.matrix(pca_data_active), eigcrit = 1, convcrit = .001)
round(print(assumptions$KMO), 3)

# Positive determinant of the correlation or variance-covariance matrices
round(det(pca_Cor), 3)

# Sphericity or the existence of the identity matrix
bartlettTest <- cortest.bartlett(pca_Cor, n = 54)
round(bartlettTest$p.value, 3)

# Plot Results

res.pca <- prcomp(pca_data_active, scale = TRUE)

groups <- as.factor(pca_data$Dmn_R_M)

fviz_pca_biplot(res.pca,
                col.ind = groups, # color by groups
                addEllipses = TRUE, # Concentration ellipses
                ellipse.type = "confidence",
                legend.title = "Raw Material",
                repel = TRUE,
                label = "var",
                pointsize = 3
) +
  theme(text = element_text(size=15), plot.title = element_text(lineheight=.8, face="bold"), plot.margin=unit(c(1,1,1,1),"cm"))

loadings_table <- round(as.data.frame(res.pca$rotation[,1:2]), 3)
